---
title: "Tutorial 00: Setup"
subtitle: "Install R, RStudio, and Quarto, then run your first interactive notebook"
format:
  html:
    toc: true
    code-copy: true
execute:
  eval: false
---

This tutorial gets your computer ready for the rest of the series. There is
no biology or single-cell data here, just installing the tools and
confirming, interactively, that they work together. By the end you will have
R, RStudio, Quarto, and Git installed, this repository cloned to your
computer, a handful of commonly used R packages available, and you will have
run a code chunk inside RStudio and seen its output.

Do this once, before Tutorial 01.

## What you'll have when you're done

- R and RStudio installed and opening correctly.
- Quarto installed (or confirmed already bundled with your RStudio).
- Git installed, and this repository cloned to your computer as an RStudio
  project.
- A small set of R packages used across these tutorials.
- A `.qmd` notebook you have opened in RStudio, run interactively chunk by
  chunk, and seen produce output (a table and a plot).

## Before you start

You need administrator rights on your computer (to install software), an
internet connection, and about 15–20 minutes. Instructions are given for
Windows and macOS separately; follow the column for your operating system.

## Step 1 — Install R

R is the language; you'll install RStudio next as the editor you actually
work in.

::: {.panel-tabset}

## Windows

1. Go to the [CRAN download page for Windows](https://cran.r-project.org/bin/windows/base/).
2. Click the top link (e.g. "Download R-4.x.x for Windows").
3. Run the downloaded `.exe` installer, accepting the default options.
4. Also install [Rtools](https://cran.r-project.org/bin/windows/Rtools/),
   matching your R version. Rtools provides the compiler some R packages need
   to install from source. It is a separate installer from R itself.

## macOS

1. Go to the [CRAN download page for macOS](https://cran.r-project.org/bin/macosx/).
2. Download the `.pkg` file matching your Mac: the `arm64` build for Apple
   Silicon (M1/M2/M3/M4), the `x86_64` build for an older Intel Mac. If
   you're not sure which chip you have, check via the Apple menu > **About
   This Mac**.
3. Run the downloaded `.pkg` installer, accepting the default options.
4. Install the Xcode command line tools, which some R packages need to
   compile from source. Open **Terminal** and run:

   ```sh
   xcode-select --install
   ```

   If a dialog says they're already installed, you're set.

:::

## Step 2 — Install RStudio

RStudio is the desktop application you'll use to write and run code, view
plots, and manage files.

::: {.panel-tabset}

## Windows

1. Go to the [RStudio Desktop download page](https://posit.co/download/rstudio-desktop/).
2. Download the Windows installer and run it, accepting the default options.

## macOS

1. Go to the [RStudio Desktop download page](https://posit.co/download/rstudio-desktop/).
2. Download the macOS installer (`.dmg`), open it, and drag RStudio into
   **Applications**.

:::

Open RStudio once it's installed. It should start without errors and show
panes for the console, environment, files, and plots. If RStudio reports it
cannot find an R installation, revisit Step 1.

## Step 3 — Install Quarto

These tutorials are written in Quarto (`.qmd`) files. Recent versions of
RStudio (2022.07 and later) bundle a copy of Quarto, so you may already have
it. Check by opening the **Terminal** tab inside RStudio (not your system
terminal) and running:

```sh
quarto --version
```

If that prints a version number, you're done with this step. If it says the
command is not found, install Quarto directly from the
[Quarto download page](https://quarto.org/docs/get-started/) for your
operating system, then restart RStudio and check again.

## Step 4 — Install Git and clone this repository

Git is the version control tool this repository is stored in. Installing it
lets RStudio pull down a copy of all the tutorials, rather than you
downloading files one at a time.

::: {.panel-tabset}

## Windows

1. Go to the [Git for Windows download page](https://git-scm.com/download/win).
2. Download and run the installer, accepting the default options throughout.

## macOS

Installing the Xcode command line tools in Step 1 already installed Git.
Confirm it's available by opening **Terminal** and running:

```sh
git --version
```

If that prints a version number, you're done. If not, install Git from the
[Git downloads page](https://git-scm.com/download/mac).

:::

Now clone the repository using RStudio's built-in Git integration:

1. In RStudio, go to **File > New Project... > Version Control > Git**.
2. In "Repository URL", paste:
   `https://github.com/tomszar/single-cell_tutorials.git`
3. Choose a location on your computer for the project to live, then click
   **Create Project**.

RStudio will download the repository and reopen itself with the project
loaded, and you'll see the tutorial folders (`tutorials/00-setup`,
`tutorials/01-read-h5ad`, ...) in the **Files** pane.

::: {.callout-note title="Just want to read the tutorials first?"}
You don't need to clone anything to read them. The rendered site is published
at <https://tomszar.github.io/single-cell_tutorials/> and always reflects the
latest version on GitHub. Cloning is only required once you want to run the
code chunks yourself.
:::

::: {.callout-warning title="RStudio's New Project dialog doesn't show a Git option"}
This means RStudio can't find your Git installation. Restart RStudio after
installing Git (Step 4 above) and try again. If it's still missing, check
**Tools > Global Options > Git/SVN** and confirm a path to the Git
executable is listed.
:::

## Step 5 — Install the basic R packages

Open RStudio, and in the **Console** pane run the following. This installs
[`tidyverse`](https://www.tidyverse.org/) (data wrangling and plotting, used
throughout this series for basic manipulation and visualization) and
`BiocManager` (the installer for Bioconductor packages, which later
tutorials use for single-cell-specific tools).

```{r}
install.packages(c("tidyverse", "BiocManager"))
```

This step downloads and compiles packages, so it can take a few minutes the
first time. Package-specific requirements for each tutorial (e.g. `schard`,
`SingleCellExperiment`) are installed separately within that tutorial; this
step only covers what's shared by all of them.

::: {.callout-warning title="Installation fails with a compilation error"}
On macOS, confirm the Xcode command line tools are installed (Step 1). On
Windows, confirm Rtools is installed and matches your R version. Restart
RStudio after installing either and try again.
:::

## Step 6 — Run a notebook interactively

This is the check that everything is wired together correctly: R, RStudio,
and Quarto's ability to run code chunks interactively.

1. In the project you cloned in Step 4, use the **Files** pane to open
   `tutorials/00-setup/index.qmd` (if the project isn't already open, use
   **File > Open Project...** and select the `.Rproj` file in the folder you
   cloned into).
2. Scroll down to the code chunk below. Click the green "play" arrow in the
   top-right corner of the chunk (or place your cursor inside it and press
   **Cmd+Enter** on macOS / **Ctrl+Enter** on Windows) to run just that
   chunk.

```{r}
library(tidyverse)

summary(mtcars)

ggplot(mtcars, aes(x = wt, y = mpg)) +
  geom_point() +
  labs(
    title = "Fuel efficiency vs. weight",
    x = "Weight (1000 lbs)",
    y = "Miles per gallon"
  )
```

If it worked, a table summary appears first, followed by a scatter plot,
both shown inline underneath the chunk (not in a separate window). That
inline, run-as-you-go behavior is what every later tutorial relies on:
you'll run one chunk, look at the result, then decide what to do next,
rather than running a whole script at once.

## If something breaks

::: {.callout-warning title="RStudio can't find R"}
Reinstall R (Step 1), then fully quit and reopen RStudio. On Windows, check
**Tools > Global Options > General** and confirm an R version is listed under
"R version".
:::

::: {.callout-warning title="`quarto --version` is not found"}
Install Quarto directly from the [Quarto download
page](https://quarto.org/docs/get-started/), then restart RStudio.
:::

::: {.callout-warning title="The chunk runs but no plot appears"}
Confirm the chunk actually loaded `tidyverse` without an error above the
plot; scroll up and check for a red error message. If `tidyverse` failed to
install in Step 5, reinstall it and re-run the chunk.
:::

::: {.callout-warning title="Package installation asks to compile from source and hangs or fails"}
Answer "no" if prompted to compile from source and let it use the
pre-built binary instead. If it still fails, revisit the compiler
requirements in Step 1 (Rtools on Windows, Xcode command line tools on
macOS).
:::

## Where this goes next

With R, RStudio, and Quarto confirmed working, you're ready for [Tutorial 01:
Read an H5AD file](../01-read-h5ad/index.qmd), where you'll load real
single-cell data from the SEA-AD atlas into R.