---
title: "Ordering cells along a trajectory"
subtitle: "Pseudotime through the oligodendrocyte lineage of one donor"
author:
  - name: "IPRE Winter 2026 · MOTCO · research line IPI-26-1206"
format:
  html:
    toc: true
    code-copy: true
execute:
  eval: false
---

The [pseudobulk track](../03a-pseudobulk/index.qmd) asked a question about
*groups*: collapse each donor to one profile per cell type and compare piles of
people. This tutorial asks the opposite kind of question—one about a *continuous
process inside a single cell type*. Some cell populations are not discrete
resting states but snapshots of cells caught partway through a gradual change.
The clearest example in brain tissue is the **oligodendrocyte lineage**:
oligodendrocyte precursor cells (**OPCs**) mature into the myelinating
**oligodendrocytes** that insulate axons, and a single snapshot of tissue
catches cells all along that path.

There is no clock in the data telling you how far along each cell is. What we can
recover instead is **pseudotime**: an ordering of cells along the progression,
built purely from how their expression profiles change. This tutorial fits that
ordering with [`slingshot`](https://bioconductor.org/packages/slingshot/), reads
off which genes turn on and off along it (myelin genes should rise; OPC genes
should fall), and—because pseudotime has no inherent direction—makes you supply
the biology that says which end is the beginning.

We stay on the single donor from tutorials 01–02 (`H20.33.001`), reusing the
preprocessed object you already saved, so the whole thing runs in a class
session.

## Learning objectives

By the end of this tutorial you should be able to:

- Say when a **trajectory / pseudotime** question is the right one—a continuous
  process within a lineage—versus a discrete group comparison.
- Subset a lineage from a preprocessed object and **re-embed it on its own**, and
  explain why the global UMAP is the wrong space for resolving within-lineage
  structure.
- Fit a pseudotime with `slingshot`: cluster the lineage, choose a **start
  cluster from marker genes**, and extract the ordering.
- Find genes that change along pseudotime and read them as a biological program.
- Explain why pseudotime's **direction is imposed, not discovered**, and defend
  the direction you chose.

**Prerequisites:** tutorials 00–02. This tutorial reloads the object you saved
at the end of [tutorial 02](../02-preprocessing/index.qmd)
(`outputs/sce_preprocessed.rds`)—no new download, and no need to redo QC or
normalization. It assumes you are comfortable with the `SingleCellExperiment`
container and with `reducedDims()`.

**Expected outcome:** a lineage object (OPCs + oligodendrocytes) carrying a
`pseudotime` value per cell and a ranked list of genes that vary along it, saved
to `outputs/`—plus a clear sense of which parts of that result you can defend and
which are artifacts of forcing a line through the data.

## Step 1 — Install the R packages

Everything except the trajectory fitter itself is already in hand from tutorial
02 (`SingleCellExperiment`, `scran`/`scater`/`scuttle`, and `bluster`, which
`scran` pulls in for clustering). The one new tool is `slingshot`.

```{r}
if (!requireNamespace("BiocManager", quietly = TRUE)) {
  install.packages("BiocManager")
}
BiocManager::install("slingshot")
```

`BiocManager` will pull in `slingshot`'s own dependencies (including
`TrajectoryUtils` and `DelayedMatrixStats`) automatically. If you are on the Nix
dev shell, `slingshot` is already built—you do not need to run
`BiocManager::install`.

## Step 2 — What pseudotime is, and what it is not

A trajectory method assumes your cells sample a **continuous transition**. Given
their expression profiles, it lays down a smooth path through the point cloud and
gives each cell a number—its **pseudotime**—measuring how far along that path it
sits. Cells with similar profiles get similar pseudotimes; cells at the two ends
get the extreme values.

Two cautions define the whole exercise:

- **Pseudotime is not real time.** It is a *distance along a path in expression
  space*, not minutes or days. Two cells with the same pseudotime are
  transcriptionally similar; that is all the number promises.
- **The method never tells you which end is the start.** A path has two ends; the
  algorithm fits the same curve whichever way you read it. Calling one end "the
  beginning" is a *biological* claim you make—here, that OPCs precede mature
  oligodendrocytes—and you justify it with known markers, not with the trajectory
  itself.

That second point is where your judgment enters, and it is deliberately left to
you in Step 6.

::: {.callout-note title="Continuum vs. clusters—pick the right question"}
Trajectory analysis only makes sense when the biology is a *gradient*. If your
populations are genuinely discrete states with no cells in between, forcing a
curve through them invents an ordering that means nothing. The oligodendrocyte
lineage is a good trajectory target because maturation is gradual and the
intermediate cells exist. Before you fit a trajectory to any population, ask
first whether a continuum is even the right model for it.
:::

## Step 3 — Reload the preprocessed donor

Set R's working directory to this tutorial's folder (**Session > Set Working
Directory > To Source File Location** in RStudio) and reload the object from
tutorial 02. It already carries `logcounts`, QC-filtered cells, the `Subclass`
labels, and `PCA`/`UMAP` in `reducedDims()`.

```{r}
library(SingleCellExperiment)
library(scater)

sce <- readRDS("../../outputs/sce_preprocessed.rds")
sce
table(sce$Subclass)
```

If that file doesn't exist, go back and run tutorial 02 through its final save
step—this tutorial does not recompute it.

## Step 4 — Subset the lineage

The oligodendrocyte lineage is two `Subclass` labels: `OPC` (the precursors) and
`Oligodendrocyte` (the mature cells). Keep only those.

```{r}
lineage <- sce[, sce$Subclass %in% c("OPC", "Oligodendrocyte")]
table(lineage$Subclass)
```

You should have a few thousand cells—many mature oligodendrocytes and a smaller
pool of OPCs. That imbalance is expected: in cortex, most cells of this lineage
have already matured.

::: {.callout-important title="Why re-embed instead of reusing the global map?"}
The `PCA` and `UMAP` in `sce` were computed across *all* cell types. Their axes
are dominated by the huge differences *between* neurons, astrocytes,
oligodendrocytes, and so on—the coarse structure that separates cell types. That
is exactly the wrong resolution for a trajectory, which lives in the *fine*
variation *within* one lineage. So we recompute the embedding on just these
cells, letting the axes capture OPC-to-oligodendrocyte variation instead. This is
a general rule: **a focused question needs a focused embedding.**
:::

## Step 5 — Re-embed the lineage on its own

Redo feature selection and dimensionality reduction—the same operations as
tutorial 02, but now on the lineage subset so the axes describe *this* lineage's
variation. Set seeds so the embedding is reproducible.

```{r}
library(scran)

# Highly variable genes within the lineage (not the whole donor).
set.seed(1)
dec <- modelGeneVar(lineage)
hvg <- getTopHVGs(dec, n = 2000)

# PCA on those genes, then UMAP for visualization.
set.seed(1)
lineage <- runPCA(lineage, subset_row = hvg, ncomponents = 30)
set.seed(1)
lineage <- runUMAP(lineage, dimred = "PCA", n_dimred = 1:20)

plotReducedDim(lineage, "UMAP", colour_by = "Subclass")
```

On the new UMAP the OPCs should sit apart from the oligodendrocytes, with the
oligodendrocytes themselves spread out rather than collapsed to a point—that
spread is the maturation variation the trajectory will order.

## Step 6 — Cluster the lineage and choose where it starts

`slingshot` builds its path in two stages: it first connects **clusters** of
cells with a minimum spanning tree to get the branching skeleton, then smooths a
curve through them. So it needs cluster labels. We cluster on the PCA space (not
the UMAP—UMAP distances are for looking, not computing).

```{r}
library(bluster)

set.seed(2)
lineage$cluster <- clusterCells(
  lineage,
  use.dimred = "PCA",
  BLUSPARAM = NNGraphParam(k = 20, cluster.fun = "louvain")
)
table(lineage$cluster, lineage$Subclass)
```

The cross-tabulation shows which clusters are OPCs and which are
oligodendrocytes. Now the biological decision the algorithm can't make for you:
**which cluster is the start?** OPCs are the precursors, and they are marked by
genes like `PDGFRA` (the canonical OPC marker), `CSPG4`, and `PTPRZ1`. Look at
where those are expressed.

```{r}
# Mean expression of an OPC marker per cluster: the start cluster is the one
# where it is high.
round(tapply(logcounts(lineage)["PDGFRA", ], lineage$cluster, mean), 2)
```

::: {.callout-tip title="Your decision: which end is the beginning"}
We pick the start cluster programmatically as the one with the highest mean
`PDGFRA`, so the tutorial proceeds—but this is a scientific call, not a
mechanical one. Confirm it with a *second* OPC marker (`CSPG4`, `PTPRZ1`) and a
mature-oligodendrocyte marker (`PLP1`, `MOBP`, `MOG`) that should show the
*opposite* pattern. If you cannot tell the two ends apart from markers, you
cannot defend a direction—and the pseudotime you get is only an ordering, not a
progression. Which genes convince you, and would they convince a skeptic?

```{r}
start_cluster <- names(which.max(
  tapply(logcounts(lineage)["PDGFRA", ], lineage$cluster, mean)
))
start_cluster
```
:::

## Step 7 — Fit the trajectory with `slingshot`

Give `slingshot` the cluster labels, the space to work in (`PCA`), and the start
cluster. It returns the same object with the trajectory attached; `slingPseudotime`
pulls out the per-cell ordering.

```{r}
library(slingshot)

set.seed(3)
lineage <- slingshot(
  lineage,
  clusterLabels = "cluster",
  reducedDim = "PCA",
  start.clus = start_cluster
)

# The lineage(s) slingshot found, as ordered sequences of clusters:
slingLineages(SlingshotDataSet(lineage))

# Store pseudotime as a per-cell column. With a single, unbranched lineage
# there is one pseudotime vector.
lineage$pseudotime <- slingPseudotime(lineage)[, 1]
summary(lineage$pseudotime)
```

For this lineage you should get a **single** path from the OPC cluster through
the oligodendrocyte clusters—no branches. If `slingshot` reports more than one
lineage, it found a branch point (see the troubleshooting note at the end); for
now we work with the first path.

::: {.callout-note title="Why fit in PCA space but look in UMAP?"}
We fit the trajectory in `PCA` because those coordinates preserve real distances
between cells; UMAP deliberately distorts distances to make a readable picture, so
a curve fit there would chase the distortion. We *visualize* on UMAP in the next
step because it is easier to read. Fit where the geometry is honest; look where
it is legible.
:::

## Step 8 — Visualize the ordering

Color the UMAP by pseudotime: it should run smoothly from the OPC end to the
mature end.

```{r}
plotReducedDim(lineage, "UMAP", colour_by = "pseudotime")
```

To draw the fitted path itself on the UMAP, project the curve into that space
with `embedCurves` and overlay it.

```{r}
curve <- slingCurves(embedCurves(lineage, "UMAP"))[[1]]

plot(reducedDim(lineage, "UMAP"),
     col = ifelse(lineage$Subclass == "OPC", "orange", "grey70"),
     pch = 16, cex = 0.4, xlab = "UMAP 1", ylab = "UMAP 2")
lines(curve, lwd = 3)
```

::: {.callout-warning title="Read the gap honestly"}
You will likely see the OPCs sitting as an island, separated from the
oligodendrocytes by a stretch with few cells, and the curve drawn straight across
that gap. `slingshot` will *always* connect the clusters you gave it—it cannot
know the gap is real. A sparse region like this can mean the intermediate cells
are genuinely rare (this maturation step is fast), or that they were lost in
dissociation or QC. The pseudotime values *across* the gap are interpolation, not
observation. Note where your trajectory is supported by dense cells and where it
is bridging empty space.
:::

## Step 9 — Which genes change along pseudotime?

The point of the ordering is the biology it exposes: genes whose expression rises
or falls as cells progress. The quick, transparent way to find them is to
**correlate each gene's expression with pseudotime** across cells. A strong
positive correlation means the gene switches on toward the mature end; a strong
negative one means it switches off.

```{r}
# Work on the highly variable genes and cells with a defined pseudotime.
ok   <- !is.na(lineage$pseudotime)
expr <- as.matrix(logcounts(lineage)[hvg, ok])
pt   <- lineage$pseudotime[ok]

# Spearman correlation is rank-based, so it catches monotone rise/fall without
# assuming the change is linear.
cors <- apply(expr, 1, function(g) cor(g, pt, method = "spearman"))
cors <- sort(cors[is.finite(cors)])

# Genes that fall along the trajectory (high early, in OPCs):
head(cors, 15)

# Genes that rise along the trajectory (high late, in mature oligodendrocytes):
tail(cors, 15)
```

Read the two lists as biology. The genes that **fall** should be OPC identity
genes (`PDGFRA`, `CSPG4`, `PTPRZ1`, `VCAN`); the genes that **rise** should
include the **myelin program**—`PLP1`, `MOBP`, `MBP`, `MOG`, `OPALIN`—the
proteins a maturing oligodendrocyte makes to wrap axons. If that is what you see,
the ordering is tracking real maturation biology, not noise.

```{r}
# Check specific markers directly, rather than trusting the top of a list.
# Correlate straight from the full matrix so this works for any marker, not
# only the 2000 HVGs the discovery lists above were restricted to.
markers <- c("PDGFRA", "CSPG4", "PTPRZ1", "MOBP", "PLP1", "MBP", "MOG", "OPALIN")
sapply(markers, function(g) cor(logcounts(lineage)[g, ok], pt, method = "spearman"))
```

Plot a couple of them against pseudotime to see the switch, not just its
correlation.

```{r}
df <- data.frame(
  pseudotime = lineage$pseudotime,
  PDGFRA = logcounts(lineage)["PDGFRA", ],
  PLP1   = logcounts(lineage)["PLP1", ]
)

ggplot2::ggplot(df, ggplot2::aes(pseudotime, PDGFRA)) +
  ggplot2::geom_point(alpha = 0.2) +
  ggplot2::geom_smooth() +
  ggplot2::ggtitle("PDGFRA (OPC marker) falls along pseudotime")

ggplot2::ggplot(df, ggplot2::aes(pseudotime, PLP1)) +
  ggplot2::geom_point(alpha = 0.2) +
  ggplot2::geom_smooth() +
  ggplot2::ggtitle("PLP1 (myelin) rises along pseudotime")
```

::: {.callout-note title="Correlations here are modest, and that's honest"}
Don't expect correlations near 1. Most oligodendrocytes are already mature, so the
big transcriptional jump is the OPC-to-oligodendrocyte switch at the start;
across the mature cells expression is comparatively flat. A correlation of ~0.4–0.5
for a real marker is a genuine, interpretable signal at this scale, not a weak
result. The *ranked pattern*—OPC genes at one end, myelin genes at the other—is
the thing to trust.
:::

## Step 10 — Save the trajectory

Save the lineage object with its pseudotime so you can reload it without
recomputing. `outputs/` is git-ignored.

```{r}
saveRDS(lineage, "../../outputs/trajectory_oligo_H20.33.001.rds")
```

## Step 11 — Your frontier {#sec-frontier}

You now have a working trajectory. Everything above was scaffolding; the open
questions below are where the real analysis is—and several have no tidy answer.

::: {.callout-tip title="Required — stress-test the trajectory you built"}
A trajectory method *always* returns a trajectory. Your job is to decide whether
to believe this one.

1. **Defend the direction.** You chose the OPC end as the start from `PDGFRA`.
   Build the full marker case: list the genes that convince you OPCs come first
   and mature oligodendrocytes come later, and say what you'd expect to see if you
   were wrong.
2. **Probe the choices.** Re-run from Step 5 changing one thing at a time—the
   number of PCs (`n_dimred`), the clustering resolution (`k` in `NNGraphParam`),
   the HVG count. Does the *ordering of markers* along pseudotime survive, or does
   the story change with the knobs? A result that only appears at one setting is
   not a result.
3. **Interrogate the gap.** Look again at the sparse region the curve bridges
   (Step 8). Is the maturation continuum really sampled, or is `slingshot` drawing
   a line between two disconnected clouds? What would you need to see to call this
   a genuine continuum?
:::

::: {.callout-tip title="Optional / advanced — does maturation shift with pathology?"}
This is the question the SEA-AD data is really for, and it is genuinely open.
Oligodendrocyte and myelin loss is reported in Alzheimer's disease—so does the
*maturation trajectory itself* differ between low- and high-pathology donors?
Are proportionally fewer cells at the mature end? Does the myelin program turn on
at a different point?

This is a **scale-up**, exactly parallel to the pseudobulk track's cohort step
([03a, Step 8](../03a-pseudobulk/index.qmd#sec-scaleup)). Our single donor is one
pathology grade (`H20.33.001` is ADNC `Low`), so within this notebook there is
nothing to compare against. To go further you would:

- run tutorials 01–02 and this lineage subset for **several donors** spanning the
  ADNC range (you choose the cohort and its confounder balance, as in 03a);
- fit a trajectory per donor—or integrate donors and fit one shared trajectory,
  which raises its own batch-correction questions;
- compare the **distribution of pseudotime** (or the position where myelin genes
  switch on) across pathology groups.

There is no guarantee of a difference, and a null result is a real finding. This
is a research question, not an exercise with an answer key.
:::

::: {.callout-note title="Optional — a principled gene test with `tradeSeq`"}
The Spearman correlation in Step 9 is a fast first pass. The purpose-built tool,
[`tradeSeq`](https://bioconductor.org/packages/tradeSeq/), fits a smooth
(generalized additive) model of counts along pseudotime per gene and tests for
significant change, which also handles non-monotone patterns and branches.
It is heavier to run; reach for it once you trust the trajectory and want
statistics on the gene dynamics.
:::

## What you produced

- A lineage object (OPCs + oligodendrocytes from `H20.33.001`), re-embedded on
  its own, carrying a `pseudotime` per cell, saved to `outputs/`.
- A ranked list of genes that vary along the ordering, with OPC identity genes at
  one end and the myelin program at the other.
- A judgment about which parts of that trajectory are supported by data and which
  are the method interpolating across gaps.

## Where this connects

- The [pseudobulk track](../03a-pseudobulk/index.qmd) and its differential
  expression sequel treat the **donor** as the unit and compare *groups*. This
  track treats the **cell** as the unit and orders a *continuum*. Same data, two
  fundamentally different questions—knowing which one your biology calls for is
  the point.
- Both tracks start from the shared preprocessed object built in
  [tutorial 02](../02-preprocessing/index.qmd).

## If something breaks

::: {.callout-warning title="`readRDS` can't find `sce_preprocessed.rds`"}
This tutorial reuses tutorial 02's output. Run
[tutorial 02](../02-preprocessing/index.qmd) through its final save step first;
the object is written to the repo-level `outputs/` folder, reached from here as
`../../outputs/sce_preprocessed.rds`.
:::

::: {.callout-warning title="`slingshot` reports more than one lineage"}
More than one entry from `slingLineages` means it placed a branch point—your
clustering split the mature end into paths it couldn't merge. Lower the clustering
resolution (raise `k` in `NNGraphParam` to get fewer, coarser clusters) so the
oligodendrocytes stay one group, then refit. There is one pseudotime column per
lineage; `slingPseudotime(lineage)` shows them all.
:::

::: {.callout-warning title="The start cluster looks wrong (markers disagree)"}
If `PDGFRA` is not clearly highest in one cluster, your clusters may be too fine
or too coarse to isolate the OPCs. Check `table(lineage$cluster,
lineage$Subclass)`: you want a cluster that is mostly `OPC`. Adjust `k` until one
cluster captures the OPCs, then set `start.clus` to it.
:::

::: {.callout-warning title="`embedCurves` or plotting fails"}
`embedCurves` needs the target reduced dimension to exist—confirm
`reducedDimNames(lineage)` still lists `"UMAP"`. If you only need the ordering and
not the drawn curve, skip this; the `pseudotime` column is already attached from
Step 7.
:::
