---
title: "Preprocessing single-cell data"
subtitle: "QC, normalization, feature selection, and dimensionality reduction on one donor"
author:
  - name: "IPRE Winter 2026 · MOTCO · research line IPI-26-1206"
format:
  html:
    toc: true
    code-copy: true
execute:
  eval: false
---

In [tutorial 01](../01-read-h5ad/index.qmd) you loaded one SEA-AD donor into a
`SingleCellExperiment` and found where the cell-type labels, sex, and pathology
score live. Quality control and normalization both start from **raw counts**: the number of
times each gene was detected in each nucleus, straight from the sequencer. Raw
counts are not yet something you can compare cell to cell, because they are
dominated by technical noise—some droplets captured more RNA than others, some
are empty or damaged, and a handful of highly expressed genes drown out the
rest. SEA-AD's files actually ship *two* versions of the matrix: a convenience
copy that is already normalized, and the raw counts stored alongside it. Step 2
shows how to tell them apart and grab the raw one.

Preprocessing is the step that turns raw counts into an object you can actually
analyze. By the end you will have thrown out low-quality nuclei, put the
remaining cells on a common scale, picked the genes that carry biological
signal, and compressed the data down to a two-dimensional map you can look at.
Both project tracks—pseudobulk and trajectory—start from this object, so this
tutorial is the shared foundation for everything that follows.

We stay on the single donor from tutorial 01 (`H20.33.001`) so the whole thing
runs in a class session.

## Learning objectives

By the end of this tutorial you should be able to:

- Explain **why** single-cell counts need quality control and normalization
  before any downstream analysis.
- Compute per-cell QC metrics (library size, genes detected, % mitochondrial
  reads) and interpret their distributions.
- Filter low-quality cells with thresholds you can *defend*, and explain the
  difference between fixed and adaptive (MAD-based) cutoffs.
- Normalize counts—both the quick log-normalization and the more principled
  `scran` pooling approach—and say what each one corrects for.
- Select highly variable genes and explain why we don't just use all of them.
- Reduce dimensionality with PCA and then UMAP, set a seed, and read the
  resulting embedding.

**Prerequisites:** tutorials 00 and 01, and the donor `.h5ad` file already
downloaded to `data/sea-ad/donor_H20.33.001.h5ad` under the repository root.
This tutorial reuses that same file.

**Expected outcome:** a filtered, normalized `SingleCellExperiment` carrying a
`logcounts` assay, a stored set of highly variable genes, and `PCA` and `UMAP`
entries in `reducedDims()`—ready for either analysis track.

## Step 1 — Install the R packages

You already have `schard` and `SingleCellExperiment` from tutorial 01. The new
tools all come from Bioconductor and belong to the [`scran`/`scater`
ecosystem](https://bioconductor.org/books/release/OSCA/) built specifically for
single-cell work:

- **`scuttle`** — QC metrics and normalization helpers.
- **`scran`** — pooling-based size factors and variance modeling for feature
  selection.
- **`scater`** — PCA/UMAP wrappers and plotting.

```{r}
if (!requireNamespace("BiocManager", quietly = TRUE)) {
  install.packages("BiocManager")
}
BiocManager::install(c("scuttle", "scran", "scater", "uwot"))
```

`uwot` is the package that actually computes the UMAP; `scater` calls it for
us. If you are on the Nix dev shell, all of these are already built—you do not
need to run `install.packages`.

::: {.callout-note title="Expect deprecation warnings"}
This tutorial uses the `scran`/`scater`/`scuttle` functions taught in the
[OSCA book](https://bioconductor.org/books/release/OSCA/)—the standard learning
path for single-cell analysis in R. On very recent Bioconductor releases some of
these functions print a deprecation notice pointing to the newer `scrapper`
package. They still work exactly as described here, and the warnings are safe to
ignore for this tutorial; we stay on the well-documented ecosystem so you can
look up every step.
:::

## Step 2 — Reload the donor object and get the raw counts

Set R's working directory to this tutorial's folder (**Session > Set Working
Directory > To Source File Location** in RStudio) and load the donor file from
the shared `data/sea-ad/` folder.

```{r}
library(schard)
library(SingleCellExperiment)

h5 <- "../../data/sea-ad/donor_H20.33.001.h5ad"
sce <- schard::h5ad2sce(h5)
sce
```

`schard` loads the file's main matrix, `X`. For SEA-AD that matrix is **not raw
counts**—it is a log-normalized copy the authors precomputed for convenience.
You can see this: the values are fractional and small, not the whole numbers a
count matrix would have.

```{r}
assayNames(sce)                 # "X"
range(assay(sce, "X"))          # small, non-integer values (max ~8)
```

QC filtering and size-factor normalization both need the *raw* counts, so we
fetch them. SEA-AD stores them in a separate layer called `UMIs`, which
`schard::h5ad2Matrix` can read directly. It comes back without row and column
names, so we copy them from the object, then store it as the `counts` assay.

```{r}
counts_mat <- schard::h5ad2Matrix(h5, "layers/UMIs")
dimnames(counts_mat) <- dimnames(sce)
counts(sce) <- counts_mat

# We won't reuse SEA-AD's precomputed normalization; drop it to avoid confusion.
assay(sce, "X") <- NULL

# The counts should now be non-negative whole numbers.
counts(sce)[1:5, 1:5]
```

::: {.callout-note title="Why not just use the normalized `X`?"}
The point of this tutorial is to do QC and normalization yourself, with choices
you can see and defend—starting from someone else's normalized matrix would skip
exactly the steps you are here to learn. You also *can't* do count-based QC
(library size, mitochondrial fraction) on values that are no longer counts.
Different datasets hide their raw counts in different places (a `layers/` entry,
an `.h5ad` `raw` slot, a separate file); always check before assuming the matrix
you loaded is raw.
:::

## Step 3 — Compute per-cell QC metrics

A "cell" in this data is really one droplet's worth of RNA, and not every
droplet is a healthy cell. Three numbers tell you most of what you need:

- **Library size** (`sum`): total counts in the cell. Very low means an empty
  or barely captured droplet; extremely high can signal a doublet (two cells in
  one droplet).
- **Genes detected** (`detected`): how many distinct genes had at least one
  count. Low complexity is another empty-droplet signature.
- **Mitochondrial fraction** (`subsets_Mito_percent`): the share of counts from
  mitochondrial genes. When a cell's membrane ruptures, cytoplasmic RNA leaks
  out but mitochondrial RNA stays trapped, so a high mito fraction flags dying
  or damaged cells.

To compute the mito fraction we first need to know which rows are
mitochondrial genes. Real datasets don't flag them for you, so look at what the
row names and per-gene metadata actually contain before matching.

```{r}
head(rownames(sce))          # gene symbols in these files
colnames(rowData(sce))       # per-gene metadata columns
head(rowData(sce))
```

In the SEA-AD donor files the row names are **gene symbols** (the Ensembl IDs
live in `rowData(sce)$gene_ids`). Mitochondrial genes are the ones whose symbol
starts with `MT-`, such as `MT-ND1`, so we can match on the row names directly.
If you meet a dataset whose row names are Ensembl IDs instead, match on whichever
`rowData` column holds the symbols.

```{r}
library(scuttle)

is_mito <- grepl("^MT-", rownames(sce))
sum(is_mito)                 # 13 — the mitochondrial protein-coding genes
```

::: {.callout-tip title="Sanity check"}
If `sum(is_mito)` is `0`, you matched the wrong identifiers—the symbols are
somewhere other than where you looked. If it is in the hundreds, you probably
matched Ensembl IDs by accident. The human mitochondrial genome encodes 13
protein-coding genes, so a count of 13 is what you want here.
:::

Now compute the metrics and attach them to `colData(sce)`.

```{r}
sce <- addPerCellQCMetrics(sce, subsets = list(Mito = is_mito))

# The new per-cell columns:
colnames(colData(sce))
summary(sce$sum)
summary(sce$detected)
summary(sce$subsets_Mito_percent)
```

## Step 4 — Filter low-quality cells

Look at the distributions before you cut anything. Thresholds you can't see the
reason for are thresholds you can't defend.

```{r}
library(scater)

plotColData(sce, y = "sum") + ggplot2::scale_y_log10() +
  ggplot2::ggtitle("Library size")
plotColData(sce, y = "detected") + ggplot2::scale_y_log10() +
  ggplot2::ggtitle("Genes detected")
plotColData(sce, y = "subsets_Mito_percent") +
  ggplot2::ggtitle("Mitochondrial %")
```

### Fixed vs. adaptive thresholds

There are two ways to pick cutoffs:

- **Fixed thresholds** — e.g. "drop cells with fewer than 500 genes or more
  than 10% mito reads." Simple and transparent, but the right number depends on
  the tissue, the protocol, and sequencing depth. A cutoff that is sensible for
  blood may throw away perfectly good neurons.
- **Adaptive thresholds** — flag cells that are outliers *relative to this
  dataset*. The standard approach uses the **median absolute deviation (MAD)**:
  a cell is an outlier if it sits more than a few MADs from the median. This
  adapts to whatever the data's own scale is, which is why it is the OSCA
  book's default.

We'll use the adaptive approach with `perCellQCFilters`, which flags cells that
are outliers on *low* library size, *low* genes detected, or *high* mito
fraction—the three directions that indicate poor quality.

```{r}
qc <- perCellQCFilters(
  sce,
  sub.fields = "subsets_Mito_percent"
)

# How many cells each rule flags, and the total discarded:
colSums(as.matrix(qc))
```

`perCellQCFilters` works on the log scale for size and complexity (counts are
right-skewed) and flags high-mito cells in the upper tail. Attach its verdict,
inspect it, then subset.

```{r}
sce$discard <- qc$discard

# Compare discarded vs. kept cells to confirm we're cutting the bad tail, not
# a biologically interesting population.
plotColData(sce, x = "sum", y = "subsets_Mito_percent", colour_by = "discard") +
  ggplot2::scale_x_log10()

# Keep only the cells that passed.
sce <- sce[, !sce$discard]
dim(sce)
```

::: {.callout-note title="Optional: try fixed thresholds"}
As an exercise, define your own fixed cutoffs and compare. Do they keep the
same cells the adaptive rule kept?

```{r}
manual_discard <- sce$sum < 500 |
  sce$detected < 300 |
  sce$subsets_Mito_percent > 10
table(manual_discard)
```
:::

## Step 5 — Normalize

Two cells with the same biological state can still have very different total
counts purely because one was sequenced more deeply. Normalization removes that
technical difference so expression is comparable across cells. We then take the
log, which stabilizes variance and makes the highly expressed genes stop
dominating.

### The quick way: library-size normalization

The simplest correction divides each cell's counts by its library size (its
size factor) and log-transforms. `logNormCounts` does both and stores the
result in a new `logcounts` assay.

```{r}
sce <- logNormCounts(sce)
assayNames(sce)              # now includes "logcounts"
```

### The principled way: `scran` pooling

Library-size normalization assumes most genes are not differentially expressed
between cells, which breaks down when cells are very different from one another.
`scran` handles this by pooling many cells together to estimate size factors
stably, then deconvolving them back to per-cell factors. It first groups
roughly similar cells with `quickCluster` so the pooling stays within
comparable cells.

```{r}
library(scran)

set.seed(100)                # quickCluster uses randomness
clusters <- quickCluster(sce)
sce <- computeSumFactors(sce, clusters = clusters)

summary(sizeFactors(sce))    # should be positive, centered near 1

# Recompute logcounts using these better size factors.
sce <- logNormCounts(sce)
```

For a single, relatively homogeneous donor the two approaches give similar
results, but the pooling method is the one to reach for as datasets get more
heterogeneous. We keep its `logcounts` for the rest of the tutorial.

## Step 6 — Select highly variable genes

Most of the ~36,000 genes are either off or uniformly expressed and carry no
information about how cells differ. Feeding all of them into PCA just adds
noise. **Highly variable genes (HVGs)** are the genes whose variance across
cells exceeds what we'd expect from technical noise alone—those are where the
biological signal is.

`modelGeneVar` decomposes each gene's variance into a technical trend (variance
expected at that expression level) and the biological excess above it.
`getTopHVGs` then ranks genes by that biological component.

```{r}
dec <- modelGeneVar(sce)

# Visualize the mean-variance trend: the curve is the technical expectation,
# points well above it are the interesting genes.
plot(dec$mean, dec$total,
     xlab = "Mean log-expression", ylab = "Variance",
     main = "Mean-variance trend")
curve(metadata(dec)$trend(x), col = "blue", add = TRUE)

# Take the top 2000 genes by biological variance.
hvg <- getTopHVGs(dec, n = 2000)
length(hvg)
head(hvg)
```

2000 is a common default; more genes retain finer signal at the cost of more
noise. This is a knob worth trying both ways once the pipeline works.

## Step 7 — Reduce dimensionality

Even restricted to 2000 genes, the data lives in 2000-dimensional space. Two
reductions get us to something interpretable:

1. **PCA** finds the directions of greatest variation. The first few dozen
   principal components capture the dominant structure while filtering noise;
   downstream steps run on these instead of raw genes.
2. **UMAP** takes those PCs and lays them out in 2D for visualization. UMAP is
   for *looking*, not for quantitative claims—distances between far-apart
   clusters are not meaningful.

Both are randomized, so set a seed before each for reproducibility.

```{r}
set.seed(1234)
sce <- runPCA(sce, subset_row = hvg, ncomponents = 50)

# How much variance each PC explains—look for where the curve flattens.
pct_var <- attr(reducedDim(sce, "PCA"), "percentVar")
plot(pct_var, xlab = "PC", ylab = "% variance explained", type = "b")
```

```{r}
set.seed(1234)
sce <- runUMAP(sce, dimred = "PCA")

reducedDimNames(sce)         # "PCA" and "UMAP"
```

Now look at the embedding, colored by the cell-type labels from tutorial 01. If
preprocessing worked, cells of the same `Subclass` should land together.

```{r}
plotReducedDim(sce, "UMAP", colour_by = "Subclass")
```

::: {.callout-tip title="Read the map critically"}
Distinct, well-separated islands that match known cell types are a good sign.
If everything is one blob, or if the clusters line up with library size instead
of biology, revisit your QC and normalization—those are the usual culprits. Try
coloring by `sum` or `subsets_Mito_percent` to check that technical variation
isn't driving the layout.
:::

## Step 8 — Save the preprocessed object

Both tracks start here, so save the result. The repo keeps generated analysis
outputs in the top-level `outputs/` folder, which `.gitignore` already excludes
from version control—so the object won't get committed by accident.

```{r}
saveRDS(sce, "../../outputs/sce_preprocessed.rds")
```

The two analysis tracks reload it with
`readRDS("../../outputs/sce_preprocessed.rds")`.

## What you produced

Your `sce` now carries everything the downstream tracks need:

- a `logcounts` assay from `scran` pooling normalization,
- QC metrics in `colData(sce)` and the low-quality cells removed,
- a 2000-gene HVG set in `hvg`, and
- `PCA` and `UMAP` in `reducedDims(sce)`.

## Where this goes next

With a clean, normalized object in hand, the project tracks diverge:

- **Pseudobulk track:** repeat loading across donors, aggregate counts per
  donor × cell type, and run a donor-level differential expression comparison
  across AD pathology.
- **Trajectory track:** subset one cell type, reuse the PCA/UMAP you just
  computed, and fit a `slingshot` trajectory anchored in low-pathology cells.

## If something breaks

::: {.callout-warning title="`h5ad2Matrix` can't find `layers/UMIs`"}
The layer name is dataset-specific. List what the file contains with
`rhdf5::h5ls("../../data/sea-ad/donor_H20.33.001.h5ad")` and look under
`/layers` for the raw-count matrix, then pass that path to `h5ad2Matrix`.
:::

::: {.callout-warning title="`sum(is_mito)` is 0"}
The row names aren't the gene symbols you expected. Print `head(rownames(sce))`
and `head(rowData(sce))`, find where symbols like `MT-ND1` live, and grep on
that instead.
:::

::: {.callout-warning title="`runUMAP` cannot find uwot"}
Install it with `BiocManager::install("uwot")` (or `install.packages("uwot")`),
then rerun Step 7.
:::

::: {.callout-warning title="Almost every cell is discarded in QC"}
Adaptive thresholds assume most cells are good. If the donor file was already
QC'd upstream, the MAD rule can over-flag. Inspect the distributions in Step 4
and, if needed, switch to explicit fixed thresholds you can justify.
:::
