---
title: "Opening SEA-AD single-cell files in R"
subtitle: "Quick-start with a single donor"
author:
  - name: "IPRE Winter 2026 · MOTCO · research line IPI-26-1206"
format:
  html:
    toc: true
    code-copy: true
execute:
  eval: false
---

This tutorial walks through loading one SEA-AD donor's single-nucleus RNA-seq
file into R and looking at what is inside. Using one donor instead of the whole
cohort keeps the download and memory requirements small, so you can get a
working object in a few minutes before you split off into your own track.
Everything after the download stays in R.

SEA-AD ships its data as `.h5ad`, a format from the Python world. We read it
straight into R with [`schard`](https://github.com/cellgeni/schard), which is
pure R—no Python environment is required—and returns a
`SingleCellExperiment`, the standard Bioconductor object used by both
downstream pipelines.

## Before you start

You need R 4.2 or newer and a web browser. One donor file is small enough for
any laptop.

## Step 1 — Install the R packages

Run this once. The first two packages come from Bioconductor; `schard` comes
from GitHub.

```{r}
# Bioconductor packages
if (!requireNamespace("BiocManager", quietly = TRUE)) {
  install.packages("BiocManager")
}
BiocManager::install(c("rhdf5", "SingleCellExperiment"))

# schard: reads .h5ad in pure R, no Python needed
if (!requireNamespace("remotes", quietly = TRUE)) {
  install.packages("remotes")
}
remotes::install_github("cellgeni/schard")
```

If `schard` reports an installation error, install `rhdf5` on its own first—it
is the component that reads the HDF5 file—and then try installing `schard`
again.

## Step 2 — Download one donor's file in a web browser

The single-cell data is available from the public SEA-AD download site. No
account or command-line tool is required. The per-donor RNA-seq files are in
`PFC/RNAseq/donors_objects/`. PFC is the prefrontal cortex, Brodmann area 9,
also called DLPFC. There is one file per donor in that folder, plus a large
all-cohort file elsewhere in `RNAseq/` that you should skip for now.

1. Open the [SEA-AD public data browser](https://sea-ad-single-cell-profiling.s3.amazonaws.com/index.html).
2. Open the `PFC` folder, then `RNAseq`, then `donors_objects`.
3. Find the filename containing `H20.33.001` (its full name is
   `H20.33.001_SEAAD_DFC_RNAseq_final-nuclei.<date>.h5ad`).
4. Download that `.h5ad` file.
5. In the repository root, create a `data/sea-ad/` folder if it does not exist,
   then move the downloaded file into it and rename it `donor_H20.33.001.h5ad`.
   The full path from the repository root should be
   `data/sea-ad/donor_H20.33.001.h5ad`.

This is the shared data folder for the whole tutorial series. Later tutorials
reuse the same file from here, so you only download it once. It is already
git-ignored, so it will never be committed.

Any donor works; `H20.33.001` is just a concrete choice. If you use a different
donor, change the filename in the R code below. The other brain region, MTG
(middle temporal gyrus), has the same layout under `MTG/RNAseq/donors_objects/`.

## Step 3 — Load it into R

Set R's working directory to this tutorial's folder. In RStudio, select
**Session > Set Working Directory > To Source File Location**. The path below
then reaches the shared `data/sea-ad/` folder from here (`../../` steps up out
of `tutorials/01-read-h5ad/` to the repository root). Run the following code.

```{r}
library(schard)
library(SingleCellExperiment)

sce <- schard::h5ad2sce("../../data/sea-ad/donor_H20.33.001.h5ad")
sce
```

This prints a summary of the object. A `SingleCellExperiment` holds three
things you will reach for constantly:

- the expression matrix: `counts(sce)` or `assay(sce)`;
- the per-cell metadata: `colData(sce)`; and
- the per-gene metadata: `rowData(sce)`.

## Step 4 — Look at what you loaded

Get your bearings before doing anything else. This is also the raw material
for the variable dictionary you were asked to start.

```{r}
dim(sce)                 # genes (rows) x cells (columns)
assayNames(sce)          # which matrices are stored
head(rownames(sce))      # gene identifiers
colnames(colData(sce))   # every per-cell metadata column
```

Now find the columns that matter for the Alzheimer's work: the cell-type
label, the donor's sex, and the pathology measure.

```{r}
cd <- colData(sce)

# Cell-type labels in the SEA-AD taxonomy
table(cd$Subclass)     # broad types (e.g. Astrocyte, Oligodendrocyte)
table(cd$Supertype)    # fine-grained types

# Sex and the categorical AD pathology score
table(cd$Sex)
table(cd$`Overall AD neuropathological Change`)

# Continuous pathology axis: confirm its exact name, then inspect it
grep(
  "pseudo|neuropath",
  colnames(cd),
  value = TRUE,
  ignore.case = TRUE
)
```

Two things to notice:

1. Column names with spaces need backticks, as with
   `Overall AD neuropathological Change`. Copy the capitalization exactly as
   `colnames(colData(sce))` prints it—here the "n" in "neuropathological" is
   lowercase.
2. Because this is a single donor, `Sex` and the pathology score are the same
   for every cell. The differences you care about appear only when you compare
   donors, which is the next step in both tracks.

::: {.callout-important title="Your deliverable: a variable dictionary"}
As you explore, keep a running **variable dictionary**—your own notes on the
`colData` columns: what each one means, whether it is categorical or continuous,
and whether it describes a single *cell* or the whole *donor*. This is your
artifact, not a graded worksheet, and you will lean on it constantly: the
grouping variables, confounders, and contrasts you choose in later tutorials all
come from here. Start it now with the columns above—`Subclass`, `Supertype`,
`Sex`, `Overall AD neuropathological Change`, and the continuous pathology
score—and add to it whenever you meet a new column. Nobody hands you the "right"
set of variables; deciding which ones matter for your question is the first piece
of the analysis that is genuinely yours.
:::

## Step 5 — Subset to one cell type

To see how subsetting works, keep just one cell type. Match the label exactly
to what `table(cd$Subclass)` printed.

```{r}
astro <- sce[, sce$Subclass == "Astrocyte"]
dim(astro)
```

Microglia—the brain's resident immune cells and a focus of Alzheimer's
research—are worth finding too. In this taxonomy they are not their own
`Subclass`; they sit inside the broader `Immune` subclass, with `Supertype`
labels beginning `Micro-PVM`. You can select them either way:

```{r}
immune <- sce[, sce$Subclass == "Immune"]                 # whole immune compartment
micro  <- sce[, grepl("^Micro-PVM", sce$Supertype)]       # microglia specifically
dim(immune)
dim(micro)
```

## If something breaks

::: {.callout-warning title="The file cannot be found"}
Confirm that the file is at `data/sea-ad/donor_H20.33.001.h5ad` under the
repository root and that R's working directory is this tutorial's folder. Check
with:

```{r}
getwd()
file.exists("../../data/sea-ad/donor_H20.33.001.h5ad")
```
:::

::: {.callout-warning title="Integer-overflow warning, or `'p' must be a nondecreasing vector`"}
This can happen with very large `.h5ad` files, such as the full
1.2-million-cell cohort file, which is about 35 GB. Using one donor avoids the
problem; this is a known limitation of reading very large files this way.
:::

::: {.callout-warning title="Out of memory"}
Do not load the all-cohort file on a laptop. A one-donor file is appropriate
for this tutorial.
:::

::: {.callout-warning title="A `table()` result is empty or a column is oddly named"}
Run `colnames(colData(sce))` again and copy the column name exactly, including
spaces and capitalization.
:::

## Where this goes next

You now have a `SingleCellExperiment` for one donor and know where sex,
pathology, and cell type live in it.

Before you move on, spend a few minutes exploring on your own—this is where the
analysis choices you'll make later begin:

- Browse the full metadata with `colnames(colData(sce))` and a few `table()` or
  `summary()` calls. Which variables would you use to compare *cells* within this
  donor? Which only make sense when comparing *donors*?
- Which columns look like they could be **confounders**—things that might track
  with pathology but aren't pathology itself (sex, age, post-mortem interval, RNA
  quality)? Note them in your variable dictionary; you'll decide what to *do*
  about them when you run a real comparison.

There isn't one right answer here—the point is to start reasoning about this as
*your* dataset rather than a fixed recipe.

From here, the project tracks diverge:

- **Pseudobulk analysis:** repeat this loading process over donors, then
  perform per-donor pseudobulk aggregation.
- **Trajectory analysis:** run PCA and UMAP on your chosen cell type, then
  anchor the `slingshot` trajectory in low-pathology cells.

Bring your variable dictionary to the next check-in.
