---
title: "Building pseudobulk profiles from single cells"
subtitle: "Collapsing a donor's cells into one count profile per cell type"
author:
  - name: "IPRE Winter 2026 · MOTCO · research line IPI-26-1206"
format:
  html:
    toc: true
    code-copy: true
execute:
  eval: false
---

Tutorials [01](../01-read-h5ad/index.qmd) and [02](../02-preprocessing/index.qmd)
stayed inside a *single* donor: enough to learn the object and the preprocessing
steps, but not enough to answer the question the SEA-AD project actually cares
about: **which genes change with Alzheimer's pathology?** That is a comparison
*between* people, and comparing people is *differential expression*, which is the
next tutorial ([04](../04-pseudobulk-de/index.qmd)).

Before you can compare people, though, you need the right **unit** to compare.
The mistake, once you have hundreds of thousands of cells, is to treat every cell
as an independent data point and test gene by gene: pool all the astrocytes from
low-pathology donors, all the astrocytes from high-pathology donors, and compare
the two piles. That answer is statistically wrong (we'll see why below), and the
fix is **pseudobulk**: collapse each donor's cells of a given type into *one*
summed count profile. This tutorial builds that unit for a single donor. Stacking
many donors' profiles (the input the differential-expression tutorial needs) is
then a short loop you'll write at the end.

## Learning objectives

By the end of this tutorial you should be able to:

- Explain why the **donor is the unit of replication** in a cross-condition
  comparison, and why per-cell tests commit *pseudoreplication*: treating
  non-independent cells as independent and massively inflating significance.
- Aggregate one donor's single-cell counts into **pseudobulk** profiles, one per
  cell type, with `aggregateAcrossCells`.
- Inspect and save a per-donor pseudobulk object, keeping the data as integer
  counts.
- Describe how to scale the same step across a cohort of donors to produce the
  matrix a differential-expression test consumes.

**Prerequisites:** tutorials 00–02. This tutorial reuses the single donor you
already downloaded in tutorial 01 (`H20.33.001`); no new download needed. It
reuses the loading pattern from tutorial 01 and the QC logic from tutorial 02.

Unlike the trajectory track, pseudobulk does **not** reuse the normalized object
from tutorial 02: it works from *raw counts* (you sum counts, then let the DE
step do its own normalization).

**Expected outcome:** a pseudobulk object for one donor (rows are genes, columns
are cell types, values are summed raw UMI counts), saved to `outputs/`, ready to be
stacked with other donors in tutorial 04.

## Step 1 — Install the R packages

Everything here comes from the earlier tutorials: `schard` (loading), the
`SingleCellExperiment` container, and `scuttle` (QC and aggregation). There is no
new package to install for *generating* pseudobulk; the differential-expression
tool (`edgeR`) enters in tutorial 04.

```{r}
if (!requireNamespace("BiocManager", quietly = TRUE)) {
  install.packages("BiocManager")
}
BiocManager::install(c("SingleCellExperiment", "scuttle"))
# schard is on GitHub; install with remotes if you don't have it yet:
# remotes::install_github("cellgeni/schard")
```

If you are on the Nix dev shell, these are already built, so you do not need to run
`install.packages`.

## Step 2 — Why pseudobulk? The pseudoreplication trap

Imagine two groups of donors (low pathology and high pathology) and suppose you
have 5,000 astrocytes from each group. A per-cell test (a *t*-test or a
single-cell DE method) treats those 10,000 cells as 10,000 independent
observations. They are not. Cells from the same donor share that donor's
genotype, age, medications, RNA quality, and dissection batch; they are far more
alike than cells from different donors. Statistically, they are
**pseudoreplicates**: repeated measurements of the *same* biological unit, not new
biological units.

The consequence is severe. With thousands of correlated cells, the standard
error shrinks toward zero and almost every gene comes out "significant," even
genes that differ only because of one unusual donor. The test is answering "are
these two piles of cells different?" (which they always are) instead of "does
pathology change expression *across people*?", which is the question you meant to
ask.

**Pseudobulk** restores the right unit of analysis. For each donor, sum the
counts across all cells of one type into a single profile. Now each donor
contributes *one* number per gene, so a cohort of, say, 12 donors gives 12
independent replicates, exactly the design that bulk RNA-seq methods like `edgeR`
and `limma-voom` were built for. Summing (not averaging) keeps the data as
integer counts, so the count-based noise models still apply, and it correctly
gives more weight to donors and cell types with more cells.

That is the whole idea of this tutorial: turn one donor's tens of thousands of
cells into a compact table of **one summed profile per cell type**. Do that for
every donor, stack the tables, and you have the matrix the
differential-expression tutorial tests.

::: {.callout-note title="This is now the field's default"}
Early single-cell DE papers used per-cell tests and reported thousands of hits.
Later re-analyses ([Squair et al. 2021](https://doi.org/10.1038/s41467-021-25960-2),
[Crowell et al. 2020](https://doi.org/10.1038/s41467-020-19894-4)) showed that
pseudobulk with donor-level replication controls false positives far better.
For a cross-condition comparison with biological replicates, pseudobulk is the
default recommendation.
:::

## Step 3 — The donor and its cell types

We use the donor from tutorial 01, `H20.33.001`, already saved as
`data/sea-ad/donor_H20.33.001.h5ad`. Aggregation groups cells by **cell type**,
and SEA-AD provides a ready-made label for that: the `Subclass` column (e.g.
`Astrocyte`, `Oligodendrocyte`, `L2/3 IT`, `Pvalb`). Each `Subclass` will become
one column of the pseudobulk profile.

The two donor-level columns we also carry along, so the profile remembers *whose*
cells it summed, are `Donor ID` and `Overall AD neuropathological Change` (the
donor's ADNC pathology grade). These are constant within a donor.

## Step 4 — Load and lightly QC the donor

We load the donor, swap in the raw UMI counts (SEA-AD's `X` assay is already
log-normalized; see tutorial 02, Step 2), and apply the same count-based QC as
tutorial 02. Wrapping it in a function pays off in Step 8, when you reuse the
exact same step for every donor in a cohort.

```{r}
library(schard)
library(SingleCellExperiment)
library(scuttle)

data_dir <- "../../data/sea-ad"

load_donor <- function(id) {
  h5 <- file.path(data_dir, paste0("donor_", id, ".h5ad"))

  # Load the object and swap in the raw UMI counts (see tutorial 02, Step 2).
  sce <- schard::h5ad2sce(h5)
  counts_mat <- schard::h5ad2Matrix(h5, "layers/UMIs")
  dimnames(counts_mat) <- dimnames(sce)
  counts(sce) <- counts_mat
  assay(sce, "X") <- NULL

  # Count-based QC: drop low-quality nuclei (tutorial 02, Steps 3–4).
  is_mito <- grepl("^MT-", rownames(sce))
  sce <- addPerCellQCMetrics(sce, subsets = list(Mito = is_mito))
  discard <- perCellQCFilters(sce, sub.fields = "subsets_Mito_percent")$discard
  sce <- sce[, !discard]

  sce
}

sce <- load_donor("H20.33.001")
sce
```

Confirm the columns we group and label on are present, and that the donor-level
ones really are constant within this donor.

```{r}
table(sce$Subclass)                                  # cell types to aggregate over
unique(sce$`Donor ID`)                               # one value: the donor
unique(sce$`Overall AD neuropathological Change`)    # one value: this donor's ADNC
```

## Step 5 — Aggregate to pseudobulk

`aggregateAcrossCells` (from `scuttle`) sums the count matrix within groups of
cells. We group by **cell type** (`Subclass`); because this is a single donor,
every cell already belongs to the same donor, so the result is one summed profile
per cell type for this one donor. The donor-level columns carry through unchanged,
and the result records how many cells went into each sum (`ncells`).

```{r}
pb <- aggregateAcrossCells(
  sce,
  ids = sce$Subclass,
  use.assay.type = "counts",
  statistics = "sum"
)
pb
```

`pb` is itself a `SingleCellExperiment`: same genes as before (the rows), but now
just one column per cell type instead of tens of thousands of cells.

::: {.callout-important title="Aggregate raw counts, not normalized values"}
Pseudobulk sums **counts**, which is why tutorial 02's normalized `logcounts`
are the wrong input here: you cannot meaningfully sum log-normalized values, and
the DE step needs integer counts to model. We deliberately reloaded the raw
`UMIs` layer in `load_donor` for exactly this reason.
:::

## Step 6 — Inspect the profiles

Look at what you built: how many cells stand behind each cell type's profile, and
what the summed counts look like.

```{r}
# Cells summed into each cell-type profile, largest first.
sort(pb$ncells, decreasing = TRUE)

# The aggregation key is stored in `ids`; name the columns by cell type.
colnames(pb) <- pb$ids

# A corner of the pseudobulk matrix: genes (rows) x cell types (columns).
assay(pb, "counts")[1:5, 1:5]
```

Two things to notice. First, the columns are wildly different in depth: an
abundant type like `L2/3 IT` may sum millions of counts while a rare type sums a
few thousand. That is expected, and the DE step's normalization (tutorial 04)
handles it. Second, a profile summed from only a handful of cells is mostly
noise; when you scale up, you'll drop cell types with too few cells per donor
(a common cutoff is `ncells >= 10`).

## Step 7 — Save the per-donor profile

Save this donor's pseudobulk so the next tutorial (and your own scale-up) can
reuse it without recomputing. `outputs/` is git-ignored.

```{r}
dir.create("../../outputs", showWarnings = FALSE)
saveRDS(pb, "../../outputs/pseudobulk_H20.33.001.rds")
```

## Step 8 — Scale up: from one donor to a cohort {#sec-scaleup}

Differential expression needs **many** donors (the donor is the replicate) spanning
the pathology range so there is something to compare. SEA-AD grades each donor
with **Overall AD neuropathological Change (ADNC)**: `Not AD`, `Low`,
`Intermediate`, `High`. A common two-group split is **low** (`Not AD`, `Low`)
versus **high** (`Intermediate`, `High`).

Everything you need to build the cohort matrix is already in hand: `load_donor`
works for *any* donor ID, and `aggregateAcrossCells` collapses each one the same
way. The scale-up is a loop.

::: {.callout-tip title="Exercise — build the cohort pseudobulk"}
This is your first substantial autonomous step, and the input to tutorial 04.

1. **Choose a cohort.** Assemble a set of donors that spans pathology and is
   *balanced* on the obvious confounds: sex above all, and ideally comparable age,
   post-mortem interval (PMI), and RNA quality (RIN). Donor-level metadata for
   every SEA-AD donor is in the file
   `PFC/RNAseq/SEAAD_DFC_RNAseq_final-nuclei_metadata.<date>.csv` on the
   [data browser](https://sea-ad-single-cell-profiling.s3.amazonaws.com/index.html);
   the columns you need are `Donor ID`, `Sex`, `Age at Death`, `PMI`, `RIN`, and
   `Overall AD neuropathological Change`. Download each chosen donor's `.h5ad`
   into `data/sea-ad/` exactly as in tutorial 01.
2. **Loop and aggregate.** Load each donor, aggregate it to pseudobulk, and keep
   only the small pseudobulk object, so only one full-size donor is ever in memory:

   ```{r}
   donor_ids <- c("H20.33.001", "H19.33.004", "H20.33.018" #, ... your cohort
   )

   pb_list <- lapply(donor_ids, function(id) {
     message("Aggregating ", id)
     donor_sce <- load_donor(id)
     aggregateAcrossCells(donor_sce, ids = donor_sce$Subclass,
                          use.assay.type = "counts", statistics = "sum")
   })
   ```
3. **Combine into one object**, columns = donor × cell type:

   ```{r}
   # cbind needs identical genes in identical order across donors.
   common  <- Reduce(intersect, lapply(pb_list, rownames))
   pb_list <- lapply(pb_list, function(x) x[common, ])
   pb_all  <- do.call(cbind, pb_list)
   saveRDS(pb_all, "../../outputs/pseudobulk_cohort.rds")
   ```

**Optional / advanced:** a class-sized cohort is a compromise. More donors mean
more statistical power and less sensitivity to any single unusual brain; the donor
is your replicate, so *donor count*, not cell count, is what limits the analysis.
How many donors do you think you'd need to reliably detect a real change? You'll
get an empirical feel for that in tutorial 04.
:::

## What you produced

- A per-donor pseudobulk object for `H20.33.001`: genes × cell type, summed from
  raw UMIs after per-cell QC, saved to `outputs/`.
- A reusable `load_donor` function and the recipe to stack many donors into the
  cohort matrix that differential expression consumes.

## Where this connects

- [Tutorial 04 — pseudobulk differential expression](../04-pseudobulk-de/index.qmd)
  takes the stacked cohort matrix and tests, within one cell type, which genes
  change between low- and high-pathology donors.
- The other analysis track, [trajectory](../03b-trajectory/index.qmd), asks a
  different question of the same data: *ordering* cells along a continuous process
  rather than *comparing* groups of donors. Pseudobulk's unit is the donor;
  trajectory's is the individual cell.

## If something breaks

::: {.callout-warning title="`aggregateAcrossCells` can't find `Subclass`"}
Column names must match exactly. Reprint them with
`colnames(colData(sce))` and copy the label as shown (see tutorial 01 on
backticks for names with spaces).
:::

::: {.callout-warning title="`cbind` fails during scale-up: rows/genes don't match"}
`do.call(cbind, pb_list)` needs every donor to have the same genes in the same
order. Different donor files can carry slightly different feature sets; intersect
to the common genes first, as shown in Step 8 (`common <- Reduce(intersect,
...)`), before combining.
:::

::: {.callout-warning title="A cell type has very few cells"}
A pseudobulk profile summed from a handful of nuclei is mostly noise. Check
`sort(pb$ncells)`; when you scale up, drop donor × cell-type columns below a
cutoff (e.g. `ncells >= 10`) rather than trusting them.
:::
